---
title: "10/24/2022 Lecture"
author: "Lucas Welch"
date: "2022-10-24"
output: html_document
---

Today: 
-Ecape characters 
-References
-Alternation

Last Time: 
  Wildcard '.' matches any character
  We could use [] to match SOME characters 
    ex[a-g, 0-4] match  a,b,c,d... or g OR 0,1,2,3,4
    ex(abc) parentheses group patterns 
    + pattern appears one or more times
    * zero times
    ^ zero at a time 
    
Want to match a left bracket followed by any characters and right bracket ([a])

Recall \n = new line

Use escape characters for any special symbol ex( \. \[ .) \+ \? )

Regex to match left bracket, any character, right bracket is 
\[.\] matches "a[g]" 

does not match "ag[]"

Regex: Takes a string and outputs a regex
If you want to create regex = \[.\], you need to put "\\[.\\]" in a string

Regex to match any two characters followed by the same characters(..) -> two character that matches are in stored in \1 (...) -> \2 


Regex to match abc or cde 
(abc)|(cde)

```{r}
library(tidyverse)
```
```{r}
regex ("\\[a.b\\]")
```

The string "\\[a.b\\]" will translate to regex "\[a.b\]"

```{r}
str_detect(c("[apc]", "[gca]"), "\\[a.b\\]")
```
```{r}
str_extract(c("[apc]", "[gca]"), "\\[a.b\\]")
```

#Alteration

```{r}
str_extract(c("abcde", "cde", "bigeh"), "(abc)|(cde)")
```

```{r}
str_extract_all(c("abcde", "cde", "bigeh"), "(abc)|(cde)")
```

#References

```{r}
str_extract(c("abcde", "cde", "bgeh", "abcbc"), "(..)\\1")
```

```{r}
str_replace("abcbcg", "(..)\\1", "\\1")
```

Suppose I want to replace the characters inside brackets with the characters inside the paranthesis.

```{r}
str_replace(c("abcbcg", "[a]", "[]", "\\[(.*)\\]", "\\(\\1\\)"))
```

"\\[(.*)\\]" This gets turned into regex "\[(.*)\] which means, match a left bracket followed by zero or more characters, followed by a right bracket 

Because the "zero or mroe characters  part" surrounded by paranthesis, what matches inside the paraenthesis gets assigned to \1. 


#Why use regular expressions 

Still situations where compactness is important

Regular expressions are plain text

Still precise about the pattern





