---
title: 'Homework 10 CSCI 036 Solutions'
author: 'Lucas Welch'
output: 
  html_document:
    css:  ../homework-style.css
  pdf_document:
    includes:
      in_header:  "../homework-style.tex"
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 5)
#Color Format
colFmt = function(x,color){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}
#Box Format
boxFmt = function(x){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\framebox{\\hspace*{1em} ", x, "}")
  else if(outputFormat == 'html')
    paste("<span class=\"boxed\">", x , "</span>")
  else
    x
}
library(tidyverse)
```

Due:  Friday, 2022-12-02

# Instructions

When your .Rmd file knits and is complete, please submit to the autograder in Gradescope.

\newpage

:::: {.problem data-latex=""}
Consider the following hexadecimal numbers:
\[
\texttt{20, A3, 11}
\]
These can be written as decimal numbers in a vector as:
```{r}
example01 <- c(32, 163, 17)
```

Now consider the following hexadecimal numbers:
\[
\texttt{4d, 87, A4, FF}
\]
Assign the decimal equivalents of these numbers to the vector `answer01`.


::::

:::: {.solution}

```{r}
answer01 <- c(77, 135, 164, 255)
```



::::

\newpage

:::: {.problem data-latex=""}
Consider the following hexadecimal numbers:
\[
\texttt{20, A3, 11}
\]
These can be written as binary numbers in a vector using strings:
```{r}
example02 <- c("00100000", "10100011", "00010001")
```

Find the binary equivalent of the following hexadecimal numbers:
\[
\texttt{4d, 87, A4, FF}
\]
and store them as a vector of strings in `answer02`.
::::

:::: {.solution}


```{r}
answer02 <- c("01001101", "10000111", "10100100", "11111111")
```



::::

\newpage

:::: {.problem data-latex=""}
Find the decimal equivalent of the following binary numbers:
\[
\texttt{10101010, 00000000, 11111111, 01011110}
\]
Place the values into a vector of length 4 called `answer03`.
::::

:::: {.solution data-latex=""}

```{r}
answer03 <- c(170, 0, 255, 94)
```


::::

\newpage

:::: {.problem data-latex=""}
Find the hexadecimal equivalent of the following binary numbers:
\[
\texttt{10101010, 00000000, 11111111, 01011110}
\]
Store this as a vector of strings of length 4 called `answer04`.
::::

:::: {.solution data-latex=""}

```{r}
answer04 <- c("AA", "00", "FF", "5E")
```


::::

\newpage

:::: {.problem data-latex=""}
Using [charToRaw]{class=KeywordTok}, find the ASCII values for the following characters as a vector of hexadecimal values stored as string.  Then convert these hexadecimal values to decimal values and assign the resulting vector to `answer05`.
\[
\texttt{AZaz%!}
\]
::::

:::: {.solution data-latex=""}


```{r}
charToRaw("AZaz%!")
```

```{r}
answer05 <- c(65, 90, 97, 122, 37, 33)
```




::::


\newpage

:::: {.problem data-latex=""}
What character is `U+0072`?  Assign this character as a string to `answer06`.
::::

:::: {.solution}

```{r}
answer06 <- ("r")
```


```{r}
cat("\U0072")
```







::::


\newpage

:::: {.problem data-latex=""}
How many bytes does `U+0072` require for encoding in UTF-8?  Assign this integer to `answer07`.
::::

:::: {.solution data-latex=""}

```{r}
answer07 <- 1
```



::::

\newpage

:::: {.problem data-latex=""}
Given the UTF-8 value `0xC0A6`, how many bits are used to encode the Unicode character?  Assign this integer to `answer08`.
::::

:::: {.solution data-latex=""}


```{r}
answer08 <- 11
```


::::

\newpage

:::: {.problem data-latex=""}
If some data has been labeled in a dataset, \_\_\_\_\_\_ learning is used.  Fill in the blank and assign the result as a string to `answer09`.
::::

:::: {.solution}

```{r}
answer09 <- "supervised"
```



::::

\newpage

:::: {.problem data-latex=""}
If the goal is to learn features of a dataset without any prior labeling, \_\_\_\_\_\_ learning is used.  Fill in the blank and assign the result as a string to `answer10`.
::::

:::: {.solution}

```{r}
answer10 <- "unsupervised"
  
```


::::

