---
title: 'Lab:  Strings and Regular Expressions'
author: "Mark Huber"
output: 
  html_document:
    css:  ../lab-style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 5)
options(tibble.width = 62)
```

## Packages

The packages needed for the lab today are:

* [tidyverse]{class=PackageName}  This includes the following three parts.

    * [stringr]{class=PackageName} for finding regular expressions in strings.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
```

## Detecting patterns of letters

A *regular expression*, or *regex* for short, is a way of describing patterns of characters that might appear within a string.  Let's set up an example vector of strings to try this out.

```{r}
s <- c("abba", "cdas", ".cds", "hg77", "cdcd", "cd")
```


The simplest type of regex is just a search for a fixed set of characters.  This can be created by putting the letters into a string, and putting it through the function [regex]{class=KeywordTok}

```{r}
regex("cd")
```

A regex has a lot of properties!  When placed into a function like [str_detect]{class=KeywordTok}, the regex looks at each string in the input vector of character strings, and decides [TRUE]{class=OtherTok} or [FALSE]{class=OtherTok} if the pattern appears.

```{r}
str_detect(s, regex("cd"))
```

The output indicates that the pattern was found in the second, third, fifth, and sixth entries of the vector `s`.  A vector whose components are true or false is also called a *boolean vector*.  

The [str_detect]{class=KeywordTok} function could also be used to create a new function.  For instance, the following creates a function `find_cd` that searches the input vector of character strings for the `cd` pattern.

```{r}
find_cd <- function(t) str_detect(t, regex("cd"))
find_cd(s)
```

:::: {.problem}
Write a function `answer01` that takes as input a vector of character strings, and returns as output a boolean vector that says true or false whether each component of the vector contains the letters `bb`.
::::

```{r}
answer01 <- function(t) str_detect(t, regex("bb"))
answer01(s)
```

:::: {.solution}

::::

If you forget to write the [regex]{class=KeywordTok} function, don't worry:  if functions like [str_detect]{class=KeywordTok} see only a string as input, they will apply the regex function first to turn it into a regular expression.

```{r}
find_cd2 <- function(t) str_detect(t, "cd")
find_cd2(s)
```

## Anchors away!

An *anchor* can force the pattern to be at the beginning or end of the string.  The circumflex `^` forces the pattern to be at the beginning, the dollar sign `$` forces the pattern to be at the end, and both can be used as well.

```{r}
tibble(
  s,
  `has cd at beginning` = str_detect(s, "^cd"),
  `has cd at end` = str_detect(s, "cd$"),
  `has cd at beginning and end` = str_detect(s, "^cd$"),
)
```

Note that `^cd$` means that *the same two letters cd* have to be at the beginning and end.  Having two different pairs of letters as in `cdcd` does not cut it.

:::: {.problem data-latex=""}
Write a function `answer02` that takes as input a vector of character strings, and returns true if and only if the string starts with the letters `Agent`.
::::

```{r}
answer02 <- function(t) str_detect(t, "^Agent")
answer02(s)
```

:::: {.solution data-latex=""}

::::

:::: {.problem data-latex=""}
Write a function `answer03` that takes as input a vector of character strings, and returns true if and only if the string ends with the letters `Smith`.
::::

```{r}
answer03 <- function(t) str_detect(t, "Smith$")
answer03(s)
```

:::: {.solution data-latex=""}

::::



## The ultimate wildcard:  the dot

A dot (aka period) `.` that appears in a regex will match any character.  So the regex `.c` will match any character followed by a `c`.

```{r}
tibble(
  s,
  `has character followed by a c` = str_detect(s, ".c")
)
```

Note that the actual period in `".cds"` is a character, so it matched the wildcard character `.`.  But `dc` also matched the wildcard character in `.c`, so the fifth hit the pattern as well.

:::: {.problem data-latex=""}
Create a function `answer04` that takes an input a vector of character strings, and returns a boolean vector that is true if and only if the string contains an `a` followed by any character followed by a `c`.
::::

```{r}
answer04 <- function(t) str_detect(t, "(a.c)")
answer04(s)
```

:::: {.solution data-latex=""}

::::

## Extracting patterns

To see the pattern found with the wildcard, use [str_extract]{class=KeywordTok}.

```{r}
tibble(
  s,
  `character followed by a c` = str_extract(s, ".c")
)
```

:::: {.problem data-latex=""}
Write a function `answer05` that takes as input a vector of character strings, and returns a pattern that is any character followed by a `4`.  If the pattern does not exist in the string, it should return [NA]{class=OtherTok}.
::::

```{r}
answer05 <- function(t) str_extract(t, ".4")
answer05(s)
```

:::: {.solution data-latex=""}

::::


:::: {.problem data-latex=""}
Write a function `answer06` that takes as input a vector of character strings, and returns a boolean vector that checks if the string is exactly three characters, the first of which must be an `a`, and the third must be a `c`.
::::

```{r}
answer06 <- function(t) str_detect(t, "^a.c$")
answer06(s)
```

:::: {.solution data-latex=""}

::::

## Using backslashes within strings

When a string is converted into a regex using [regex]{class=KeywordTok}, it is important to understand how to create *escape characters*.  An escape characters is a backslash `\` followed by another character.  For instance, `\.` is the escape character that matches a `.` in the target string.

But if a backslash is used within a string, then that itself becomes an escape character.  For instance, `\n` is the escape character for newline in a string, so the following prints over two lines.
```{r}
cat("This is the first line.\nThis is the second line.")
```

So to make a backslash within the regex, the escape character for backslash needs to be used within the string!  This is `\\`.  So put this in the string whenever you want a `\` to appear in the regex.

For instance, the following expression searches for the pattern of any character followed by a dot in the string.
```{r}
str_extract("srt.42", regex(".\\."))
```

:::: {.problem data-latex=""}
Write a function `answer07` that takes a vector of character strings, and for each extracts the first appearance of a dot followed by a c followed by another dot.
::::

```{r}
answer07 <- function(t) str_extract(t, "\\.c\\.")
answer07(s)
```
:::: {.solution data-latex=""}

::::

:::: {.problem data-latex=""}
Another useful escape character is the quote symbol `"`.  As with a `.`, the escape character is `\"`.  To use this as part of your string, it is useful to surround your string with quote marks `'` rather than the double quote `"`.  Also, remember that you need to use two backslashes in a string to make one backslash in a regex.  Write a function `answer08` that takes a vector of character strings, and checks that there is a quote at the beginning of a string, returning a boolean vector.
::::

```{r}
answer08 <- function(t) str_detect(t, '\\"')
answer08(s)
```


:::: {.solution data-latex=""}

::::


## Brackets

Brackets in regular expressions are another way to create wildcards.  For instance, the regex `[abc]` will matches either an `a`, a `b`, or a `c`.  Hyphens can be used to make ranges.  For instance, `[3-7]` will match digits from 3 through 7.  These can be combined.  For instance, [a-c3-7] will match the letters `a` through `c` or the digits `3` through `7`.

:::: {.problem data-latex=""}
Write a function `answer09` that takes as input a vector of strings and returns a boolean vector that is true for each component if the string contains either the digits 0 through 9 or the letters A through G.
::::
```{r}
answer09 <- function(t) str_detect(t, "[A-G0-9]")
answer09(s)
```

:::: {.solution data-latex=""}

::::

## Repeating a letter or expression

The `+` sign can be used after a symbol to repeat it one or more times, and the Kleene Star `*` can be used to repeat it zero or more times.  Parentheses can be used to group patterns or objects.

For instance, to pull out the letters between `b` and `e` including the endpoints, use:
```{r}
str_extract("abghweueriv", "b[a-z]*e")
```

:::: {.problem data-latex=""}
Write a function `answer10` that extracts from an input vector of strings the pattern of a hyphen followed by zero or more digits 0 through 9 followed by a hyphen.
::::
```{r}
answer10 <- function(t) str_extract(t, "-[0-9]*-") 
answer10(s)
```

:::: {.solution data-latex=""}

::::
