---
title: 'Lab:  Working with MySQL'
author: "Mark Huber"
output: 
  html_document:
    css:  ../lab-style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse)
#library(RSQLite)
#sales <- dbConnect(SQLite(), "data_output/sales.sqlite")
```

:::: {.problem data-latex=""}
How many tables are there in the Northwind database?  Place your answer in the integer variable `answer01`.
::::

:::: {.solution}
```{r}
answer01 <- 29
```
::::


:::: {.problem data-latex=""}
After running the above commands, how many employees are there in the company?  Assign your numerical answer to `answer02`.
::::

:::: {.solution data-latex=""}
```{r}
answer02 <- 9
```
::::

:::: {.problem data-latex=""}
How many variables are there in the relation `Employees`?  Assign the answer to `answer03`.
::::

:::: {.solution}
```{r}
answer03 <- 19
```

::::


:::: {.problem data-latex=""}
Which employee comes first when ordered by last name in alphabetical order?  Assign the last name as a string to `answer04`.
::::

:::: {.solution data-latex=""}
```{r}
answer04 <- "Buchanan"
```
::::


:::: {.problem data-latex=""}
If the non-Sales Representatives are ordered by `LastName`, what is the `Title` of the first observation?  Place this as a character string into `answer05`.
::::

:::: {.solution}
This is 
```{r}
answer05 <- "Sales Manager"
```
::::


:::: {.problem data-latex=""}
Adding to the above SQL query, order the employees by the new variable `Name` in descending order.  What is the `Name` value for the first employee in the list?  Assign this as a string to `answer06`.
::::

:::: {.solution}
```{r}
answer06 <- "Steven Buchanan"
```
::::


:::: {.problem data-latex=""}
Suppose that freight with cost over \$1000 is taxed at 12\%.  Create a query to find a table with the taxed freight cost that only contains those observations where the tax applies.  How many such observations are there?  Place the result into `answer07`.
::::

:::: {.solution}
```{r}
answer07 = 1
```
::::


:::: {.problem data-latex=""}
Create an SQL query that returns for each product (determined by `OrderID`) in the `Order Details` table, the number of times that product was ordered.  Note that for tables names like this with a space in them, you must surround the name by backticks (`) to indicate that it is all one name.  Place into `answer08` the number of times that `OrderID` number 10249 was ordered
:::::

:::: {.solution data-latex=""}
```{r}
answer08 <- 2
```
::::


:::: {.problem data-latex=""}
The `ContactName` within the `Customers` dataset holds the name of the person to contact at a particular company.  

What is the name of the contact at the company that placed order with ID number 10290?  Place this string into the variable `answer09`.
::::

:::: {.solution}
```{r}
answer09 = "Pedro Afonso"
```
::::


:::: {.problem data-latex=""}
By looking at the above report, what is the company with phone number (171) 555-7788?  Place this string in `answer10`.
::::

:::: {.solution}
```{r}
answer10 <- "Around the Horn"
```
::::